Solving the Shallow Water Equations\\ using Finite Volumes and Lax-Friedrichs

### 2018-01-13 paper/paper.tex
Straggler changes to the paper draft

\begin{abstract}
Lax-Friedrichs and Local Lax-Friedrichs are two numerical methods for solving the shallow water equations. 
1 sentence intro
2 sentences methods
1 sentence impl/results
why do people care?
\end{abstract}

@@

The CFL condition is encoded on lines 93-96.

@@

The Lax-Friedrichs flux function is implemented on lines 60--72.  It was run against two model problems. The first used a Gaussian perturbation for its initial condition with reflecting boundary conditions. The second used a breaking dam initial condition with outflow boundary conditions. Two experiments were run in order to examine the behavior of the knobs $n_{cells}$  and $\mu$ independently. These are depicted in Figures~\ref{fig:lxf_gaussian_ncells},~\ref{fig:lxf_breakingdam_ncells}, and~\ref{fig:lxf_breakingdam_mu}.

@@

\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/artificial_viscosity.png}%
        \caption{Wavespeed over the spatial domain}\label{fig:artificial_viscosity}%
    \end{center}%
\end{figure}

@@

The Local Lax-Friedrichs method does not entirely break the coupling between dissipation and $\mu$, but it strongly reduces it, as depicted in Figure~\ref{fig:llxf_breakingdam_mu}. This experiment used the exact same parameters as Figure~\ref{fig:lxf_breakingdam_mu}, substituting only the LxF kernel (as implemented in lines 60--72) with the LLxF kernel (as implemented in lines 73--86).
\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/llxf_breakingdam_mu.png}%
        \caption{Local Lax-Friedrichs, breaking dam scenario. }\label{fig:llxf_breakingdam_mu}%
    \end{center}%
\end{figure}

@@

\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_breakingdam.png}%
        \caption{Comparison of Lax-Friedrichs vs Local Lax-Friedrichs, breakin dam scenario}\label{fig:cmp_breakingdam}%
    \end{center}%
\end{figure}
\begin{figure}[h]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_gaussian.png}%
        \caption{Comparison of Lax-Friedrichs vs Local Lax-Friedrichs, gaussian perturbation scenario.}\label{fig:cmp_gaussian}%
    \end{center}%
\end{figure}

@@

@@

There are several tradeoffs associated with using an array-of-structures compared to a structure-of-arrays. In performance-critical code, this choice is driven primarily by memory access patterns, which are not considered here. The principal benefit of this representation is that it moves more of the semantics of the problem domain into the type system. On the other hand, this representation is initially unaware of the fact that $Q$ forms a vector space, resulting in duplicated, messier, and possibly less efficient code.

This problem may be resolved quickly. Methods in Julia have multiple dispatch, which means that a function is resolved to a method by examining all of the types in the signature, rather than simply the `owning' object.\cite{doi:10.1137/141000671} This means that operations such addition can be extended to custom types in a unified way. Lines 12--26 extend addition and scalar multiplication over the state space type $Q$.

Let us now consider a clean organizing principle for the code. A basic skeleton for running experiments is given on lines 38--59. The following components are allowed to vary: Flux function f, initial conditions, boundary conditions, timestep size. In traditional object-oriented programming, this decoupling is usually achieved via the Strategy Pattern. Because Julia has first-class functions (lambdas), it can be achieved much more tersely.

>>>>

\begin{abstract}
Lax-Friedrichs and Local Lax-Friedrichs are two numerical methods for solving the shallow water equations. This paper introduces both and provides a concise implementation in Julia. The Gaussian perturbation and breaking dam model problems are solved for each method over a range of parameters. The cause and effects of artificial viscosity are explored analytically and experimentally. 
\end{abstract}

@@

The CFL condition is encoded on lines 95--98.

@@

The Lax-Friedrichs flux function is implemented on lines 62--74.  It was run against two model problems. The first used a Gaussian perturbation for its initial condition with reflecting boundary conditions. The second used a breaking dam initial condition with outflow boundary conditions. Two experiments were run in order to examine the behavior of the knobs $n_{cells}$  and $\mu$ independently. These are depicted in Figures~\ref{fig:lxf_gaussian_ncells},~\ref{fig:lxf_breakingdam_ncells}, and~\ref{fig:lxf_breakingdam_mu}.

@@

\begin{figure}[t]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/artificial_viscosity.png}%
        \caption{Wavespeed over the spatial domain}\label{fig:artificial_viscosity}%
    \end{center}%
\end{figure}

\begin{figure}[t]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/llxf_breakingdam_mu.png}%
        \caption{Local Lax-Friedrichs, breaking dam scenario. }\label{fig:llxf_breakingdam_mu}%
    \end{center}%
\end{figure}

@@

The Local Lax-Friedrichs method does not entirely break the coupling between dissipation and $\mu$, but it strongly reduces it, as depicted in Figure~\ref{fig:llxf_breakingdam_mu}. This experiment used the exact same parameters as Figure~\ref{fig:lxf_breakingdam_mu}, substituting only the LxF kernel (as implemented in lines 62--74) with the LLxF kernel (as implemented in lines 75--88).

@@

@@

\begin{figure}[t]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_breakingdam.png}%
        \caption{Comparison of Lax-Friedrichs vs Local Lax-Friedrichs, breakin dam scenario}\label{fig:cmp_breakingdam}%
    \end{center}%
\end{figure}

\begin{figure}[t]%
    \begin{center}%
        \includegraphics[scale=0.6]{../plots/lxf_llxf_gaussian.png}%
        \caption{Comparison of Lax-Friedrichs vs Local Lax-Friedrichs, gaussian perturbation scenario.}\label{fig:cmp_gaussian}%
    \end{center}%
\end{figure}

@@

The choice of an array-of-structures representation is most consistent with the theory, but disregards the useful fact that $Q$ forms a vector space. This problem is readily resolved thanks to \emph{multiple dispatch}. In Julia, a function is resolved to a method by examining all of the types in the signature, rather than simply the `owning' object.\cite{doi:10.1137/141000671} This means that operations such addition can be extended to custom types in a unified way. Lines 12--26 extend addition and scalar multiplication over the state space type $Q$.

The basic finite volume skeleton is given on lines 38--61. The skeleton accepts abstract strategies for flux computations, timestep computations, initial conditions, and boundary conditions. These strategies are implemented as first-class functions rather than as an object heirarchy. Lines 91--93 demonstrate a higher-order function which acts like a constructor for timestep strategies. Lines 100--102 demonstrate \emph{currying} to convert a three-argument \verb!cfl_dt()! into a two-argument version. Lines 104--112 demonstrate using the method's \emph{closure} to store arbitrary state, in this case the $\Delta t$ calculated during the preceding function call. Finally, lines 147--148 ties everything together to run Lax-Friedrichs on a breaking dam scenario.
